Predicting the effects of character displacement on food-web dynamics

### 2019-03-07 manuscript_ECD_model.Rmd
add edits to Discussion and Acknowledgements.

My results suggest that a revision is needed for one of the criteria used to demonstrate character displacement. 
Criteria #5 in @Schluter1992 states: "Sites of sympatry and allopatry should not differ greatly in food, ...". 
In contrast, my results show that character displacement leads to predictable differences in food (resource) availability and even food-web stability.
Specifically, I found that, whenever character displacement increased the effective attack rate of consumers, it resulted in lower resource abundances and less resilient food webs. 
Character displacement always increased attack rates whenever consumers were competing for resources that occured in distinct habitats. 
This occurred under different evolutionary tradeoffs and in both simple and more realistic foraging scenarios.
Although character displacement did not always increase effective attack rates, these contingent effects only occurred under the simplest and (arguably) least realistic foraging scenario. 
My study takes advantage of past work showing that adaptive divergence in consumer attack rates is always favored under the scenarios examined here [@LawlorSmith1976; @Abrams1986].
Taken together, my results indicate that the adaptive process of character displacement often comes at the ecological cost of making food webs less resilient to perturbations.

The simplest foraging scenario examined here (linear functional response on resources occuring in the same habitat) is the foundation of mechanistic models of resource competition [@MacArthur1972; @Lawlor1976; @Abrams1986]. 
Under this scenario, it is theoretically possible that character displacement results in increased resource abundances [green lines in \ref{fig:MacArthur_Plot}].
This only occurs, however, under an evolutionary tradeoff that constrains the magnitude of character displacement.
This suggests that it would be unlikely to ever observe divergent character displacement and increases in resource abundances in a real system. 
In contrast, the evolutionary tradeoff that favors character displacement the most decreased resource abundances. 
Although data on the shape of the tradeoff in consumer foraging traits is scarce, two classic examples of character displacement, Darwin's finches and threespine stickleback, both appear to exhibit a tradeoff where extreme trait values increase the foraging rate of consumers (orange lines in \ref{fig:MacArthur_Plot}A,B). 
This is evident by the disruptive selection observed for beak size/shape in Darwin's finches [@Schluter1985] and benthic/limnetic ecotypes for stickleback [@Arnegard2014] in resource-limited environments. 
Taken together, my model predicts that even in the simplest foraging scenario, the most likely relationship we would observe in nature is for food webs where consumers have undergone character displacement to have lower total resource densites.

A general result emerging from our analysis was that, when consumers were competing for resources in distinct habitats, character displacement always supressed resource abundances. 
Consumers are generally larger, and as a result, more mobile than their resources; thus, their foraging behavior acts to couple resource dynamics across traditional habitat boundaries [@McCann2005].
An example of character displacement that matches this foraging scenario is threespine stickleback in postglacial lakes of British Columbia, Canada [@Schluter1992; @Schluter2000].
In these lakes, stickleback have undergone character displacement into limnetic and benthic species that have evolved specialized traits to forage on zooplankton and benthic invertebrates, respectively.
These resources occur in distinct zones of the lake (pelagic and littoral) and stickleback move between these zones when foraging for resources. 
The theory developed here predicts that total resource abundances will be lower in lakes where stickleback have undergone character displacement compared to nearby lakes with only a single, generalist species.
A disproporationate number of the documented cases of character displacement involve carnivores that are larger, and likely more mobile, than their resources [@Schluter2000],
Therefore, my predictions may apply to many cases of ecological character displacement.

Character displacement is far from an all-or-nothing process, but may vary in its magnitude [@Schluter2000]. 
My results suggest that the magnitude of character displacement will be negatively correlated with total resource abundances, at least when resources occur in distinct habitats (line trajectories in \ref{fig:LS_Plot}B, \ref{fig:McCann_Plot}B).
This prediction could be test either experimentally, by putting populations that vary in divergence in a common resource setting, or with field data that statistically controls for other factors affecting resources.
Note though, this prediction will only apply to comparisons within species where there is likely little difference in the shape of the tradeoff among populations.

My most striking result was that character displacement, when examined through the lens of food-web theory, generally resulted in less stable food webs. 
In fact, under a foraging scenario that incorporates more biological realism, character displacement may even result in an unstable food web. 
The mechanism underlying this destabilization is quite general, resulting from the strength of interspecific interactions increasing relative to intraspecific interactions [@Chesson2008; @Rip2011; @McCann2011]. 
The strength of interspecific interactions increases whenever consumers evolved higher net attack rates, but does not change intraspecific interactions.
This relative increase in interspecific interactions, combined with the natural oscillatory tendency of consumer-resource interactions [@Murdoch2003], creates a food-web structure that is less resilient to perturbations.

My results, combined with recent theory, suggest that the very conditions that favor phenotypic diversification also make that diversity more susceptible to local extinction.
Specifically, work by @McPeek2019 showed that character displacement is favored in environments that are either highly productive, easy to find and capture resources, or under weak abiotic stress. 
In my model, this corresponds to higher values of $K$ (productivity) or $a_ij$s (finding and capturing resources), or lower values of $m$ (abiotic stress).
Each of these corresponding changes create food webs that have a lower baseline of stability, as they increase the strength of consumer-resource interactions relative to intraspecific density dependence.
For example, increasing productivity reduces intraspecific competition in resource populations while increasing the flux of energy to consumers, resulting in the "paradox of enrichment" [@Rosenzweig1971]. 
Similarly, increased capacity to find and capture resources (as we observed in our model) or lower consumer mortality, both increase the relative strength of consumer-resource interactions, which predictably destabilize consumer-resource interactions [@Rip2011; @McCann2011].

Empirical evidence supports the theoretical notion that food webs where consumers have undergone character displacement are less resilient to perturbations. 
For example, a single species of threespine stickleback lives in hundreds of postglacial lakes of British Columbia, but the species pairs, where character displacement has happened, are only known from six lakes in four independent watersheds [@Schluter1992; @Schluter2000].
Perhaps this rarity of species pairs is a result of nonadaptive selection for more ecologically stable food-web topologies [@Borrelli2015a; @Borrelli2015b].
Indeed, it is clear that these species pair are vulnerable to perturbations, given that the species pairs have went extinct in two lakes following the introduction of nonnative species [@Hatfield2001; @Taylor2006; @Rudman2016].
The vulnerability of the stickleback system also aligns well with evidence that aquatic food webs have several properties that make them less resilient to perturbations, including higher productivity and more efficient transfer of energy to consumers [@Rip2011].
Testing this hypothesis may be quite difficult, but it could be possible with recent advances in genomics.
For example, @Feulner2018 detected genomic signatures of hybridization in sympatric whitefish species following periods of eutrophication.
Perhaps stickleback in many other postglacial lakes in British Columbia retain genomic signatures of having been a habitat specialist in the past.

My stability results contrast, but do not necessarily contradict, the notion from competition theory that character displacement contributes to species coexistence [@Lawlor1976]. 
Competition theory typically uses a different measure of stability that considers the mutual ability of consumers with different phenotypes to invade when rare [mutual invasibility, @Chesson2000].
A shortcoming of this mutual invasibility measure, however, is that it does not allow a comparison between food webs with and without a competiting consumer. 
Such comparisons are necessary for inferring the effects of ecological character displacement, a point that has been made clear in the criteria to demonstrate character displacement [@Schluter1992; @Schluter2000].
Although the addition of consumers can decrease food-web resilience in the absence of evolution [@May1973], our results are primarily driven by an eco-evolutionary feedback between consumer evolution and resource abundances.

I expect that my theoretical predictions are robust to different assumptions about the underlying evolutionary dynamics of consumer foraging traits.
This is because models that explicitly include resource dynamics inevitably show that resource competition results in divergent character displacement, regardless of whether a quantitative genetic or adaptive dynamics model is used for consumer traits [@Lawlor1976; @Taper1985]. 
A quantitative genetic model may certainly show differences in the magnitude of character displacement, but this should not affect the qualtiative relationship I observed.  
In fact, recent work by @McPeek2017 has shown that divergence in a quantitative genetics model can lead to decreased resource abundances and unstable dynamics.
I want to note though, that my conclusions may only apply to food webs with biotic resources that are nutritionally substitutable. 
It would be interesting to extend these current analyses to non-substitutable resources where convergent character displacement is expected [@Abrams1987; @Fox2008].

Competition theory suggested that character displacement was a stabilizing force in ecological communities [@Lawlor1976]. 
My results show the exact opposite, when a food-web perspective is taken.
Importantly, a food-web approach permits comparisons between communities in the presence of absence of a competitor, allowing a fair comparison of the effects of character displacement on food-web dynamics.
These results appear to be quite general, as they characterize the general foraging scenario where consumers are more mobile than their resources. 
These results also suggest that the very evolutionary process that generates biodiversity, creates an ecological scenario that makes that diversity fragile to loss. 
These results highlight how evolution, when considered in a food-web context, may actually be a destabilizing force in ecological communities. 
This may generally be true of consumer evolution, as was pointed out decades ago [@Rosenzweig1963].
These results also highlight a tension between adaptive process (evolution of consumers to increase feeding efficiency) and a non-adaptive process (food-web stability).
These results suggest that in systems with evidence of ecological character displacement, why do they persist? 
These results open up a new avenue for research on ecological character displacement, one that explores in more detail the dynamical consequences of consumer evolution, rather than just static effects that assume resources are persistent and stable. 
It also suggest that actually observing evidence of ecological character displacement may be difficult ...
It also highlights how evolution of key species in a food web may make these systems more susceptible to environmental change. 
It's been argued that "inferences about the evolution of community structure cannot be based on ecological arguments about community stability or likely coexistence" [@McPeek2019].
In contrast, our results suggest that current patterns in food-web structure may be the result of non-adaptive processes selecting for more stable food-web configurations.
Our results also call for a revision of one of the criteria of character displacement, since it is clear that this process can result in different resource dynamics.
In fact, this resource similarity hypothesis has not been tested in natural systems, although it has been qualitatively inferred, and some quantitative estimates based on physical properties of systems have been measured to try and say these habitats are similar. (NEED TO CHECK SCHLUTER 2000 BOOK, THE AM NAT STUDY SUGGESTS THAT IT IS A COMMONLY FILLED CRITERIA.)
More generally, our results highlight how variation in food-web structure can alter eco-evolutionary feedbacks within communities [@McPeek2017], that also have consequences for their resilience to environmental disturbances.
This calls for food-web ecologists to more criticaly consider the role of evolutionary processes in shaping food-web structure and dynamics.

@@

The content of this paper was greatly enhanced by numerous conversations with Seth Rudman. Sally Otto provided much help in early analyses of the mathematical model and I would not have been able to take the theory as far as I did without her guidance and encouragement. A discussion with Ben Gilbert pushed me to take a comparative approach that looked across different foraging scenarios to examine the generality of my conclusions. Discussions with Dolph Schluter pushed me to also to take a comparative approach that better set my results in the context of past theory on character displacement.

>>>>

My results show that character displacement leads to predictable differences in food (resource) availability and even food-web stability.
Specifically, I found that, whenever character displacement increased the effective attack rate of consumers, it resulted in lower resource abundances and less resilient food webs. 
Character displacement always increased attack rates whenever consumers were competing for resources that occured in distinct habitats. 
This occurred under different evolutionary tradeoffs and in both simple and more realistic foraging scenarios.
Although character displacement did not always increase effective attack rates, these contingent effects only occurred under the simplest and (arguably) least realistic foraging scenario. 
My study takes advantage of past work showing that adaptive divergence in consumer attack rates is always favored under the scenarios examined here [@LawlorSmith1976; @Abrams1986].
Taken together, my results indicate that the adaptive process of character displacement often comes at the ecological cost of making food webs less resilient to perturbations.

The ecological and evolutionary scenarios that generated the greatest divergent character displacement always resulted in decreases in resource densities. <!-- build paragraph around this topic sentence -->

The simplest foraging scenario examined here (linear functional response on resources occuring in the same habitat) is the foundation of mechanistic models of resource competition [@MacArthur1972; @Lawlor1976; @Abrams1986]. 
Under this scenario, it is theoretically possible that character displacement results in increased resource abundances [green lines in \ref{fig:MacArthur_Plot}].
This only occurs, however, under an evolutionary tradeoff that constrains the magnitude of character displacement.
This suggests that it would be unlikely to ever observe divergent character displacement and increases in resource abundances in a real system. 
In contrast, the evolutionary tradeoff that favors character displacement the most decreased resource abundances. 
Although data on the shape of the tradeoff in consumer foraging traits is scarce, two classic examples of character displacement, Darwin's finches and threespine stickleback, both appear to exhibit a tradeoff where extreme trait values increase the foraging rate of consumers (orange lines in \ref{fig:MacArthur_Plot}A,B). 
This is evident by the disruptive selection observed for beak size/shape in Darwin's finches [@Schluter1985] and benthic/limnetic ecotypes for stickleback [@Arnegard2014] in resource-limited environments. 
<!-- key point of the paragraph -->
Taken together, my model predicts that even in the simplest foraging scenario, the most likely relationship we would observe in nature is for food webs where consumers have undergone character displacement to have lower total resource densites.

A general result emerging from my analysis was that, when consumers were competing for resources in distinct habitats, character displacement always supressed resource abundances. 
<!-- show that this is common scenario -->
Consumers are generally larger, and as a result, more mobile than their resources; thus, their foraging behavior acts to couple resource dynamics across traditional habitat boundaries [@McCann2005].
An example of character displacement that matches this foraging scenario is threespine stickleback in postglacial lakes of British Columbia, Canada [@Schluter1992; @Schluter2000].
In these lakes, stickleback have undergone character displacement into limnetic and benthic species that have evolved specialized traits to forage on zooplankton and benthic invertebrates, respectively.
These resources occur in distinct zones of the lake (pelagic and littoral) and stickleback move between these zones when foraging for resources. 
<!-- good closing sentences -->
A disproporationate number of the documented cases of character displacement involve carnivores that are larger, and likely more mobile, than their resources [@Schluter2000],
Therefore, my predictions may apply to many cases of ecological character displacement.

My results suggest that a revision is needed for one of the criteria used to demonstrate character displacement. 
Criteria #5 in @Schluter1992 states: "Sites of sympatry and allopatry should not differ greatly in food, ...".

<!-- consider removing entirely. This is a prediction I could test with Seth in another paper...
Character displacement is far from an all-or-nothing process, but may vary in its magnitude [@Schluter2000]. 
My results suggest that the magnitude of character displacement will be negatively correlated with total resource abundances, at least when resources occur in distinct habitats (line trajectories in \ref{fig:LS_Plot}B, \ref{fig:McCann_Plot}B).
This prediction could be tested either experimentally, by putting populations that vary in divergence in a common resource setting, or with field data that statistically controls for other factors affecting resources.
Note though, this prediction will only apply to comparisons within species where there is likely little difference in the shape of the tradeoff among populations.
-->

My most striking result was that divergent character displacement, when examined through the lens of food-web theory, generally destabilized food-web dynamics. 
In fact, under a foraging scenario that incorporates more biological realism, character displacement may even result in an unstable food web. 
The mechanism underlying this destabilization is quite general, resulting from the strength of interspecific interactions increasing relative to intraspecific interactions [@Chesson2008; @Rip2011; @McCann2011]. 
The strength of interspecific interactions increases whenever consumers evolved higher net attack rates, but does not change intraspecific interactions.
This relative increase in interspecific interactions, combined with the natural oscillatory tendency of consumer-resource interactions [@Murdoch2003], creates a food-web structure that is less resilient to perturbations.

<!-- good, but tighten up -->
My results, combined with recent theory, suggest that the very conditions that favor phenotypic diversification also make that diversity more susceptible to local extinction.
Specifically, work by @McPeek2019 showed that character displacement is favored in environments that are either highly productive, easy to find and capture resources, or under weak abiotic stress. 
In my model, this corresponds to higher values of $K$ (productivity) or $a_{ij}$ (finding and capturing resources), or lower values of $m$ (abiotic stress).
Each of these corresponding changes create food webs that have a lower baseline of stability, as they increase the strength of consumer-resource interactions relative to intraspecific density dependence.
For example, increasing productivity reduces intraspecific competition in resource populations while increasing the flux of energy to consumers, resulting in the "paradox of enrichment" [@Rosenzweig1971]. 
Similarly, increased capacity to find and capture resources (as observed in my model) or lower consumer mortality, both increase the relative strength of consumer-resource interactions, which predictably destabilize consumer-resource interactions [@Rip2011; @McCann2011].

<!-- interesting, but tighten up -->
Empirical evidence supports the theoretical notion that food webs where consumers have undergone character displacement are less resilient to perturbations. 
For example, a single species of threespine stickleback lives in hundreds of postglacial lakes of British Columbia, but the species pairs, where character displacement has happened, are only known from six lakes in four independent watersheds [@Schluter1992; @Schluter2000].
Perhaps this rarity of species pairs is a result of nonadaptive selection for more ecologically stable food-web topologies [@Borrelli2015a; @Borrelli2015b].
Indeed, it is clear that these species pair are vulnerable to perturbations, given that the species pairs have went extinct in two lakes following the introduction of nonnative species [@Hatfield2001; @Taylor2006; @Rudman2016].
The vulnerability of the stickleback system also aligns well with evidence that aquatic food webs have several properties that make them less resilient to perturbations, including higher productivity and more efficient transfer of energy to consumers [@Rip2011].
Testing this hypothesis may be quite difficult, but it could be possible with recent advances in genomics.
For example, @Feulner2018 detected genomic signatures of hybridization in sympatric whitefish species following periods of eutrophication.
Perhaps stickleback in many other postglacial lakes in British Columbia retain genomic signatures of having been a habitat specialist in the past.

<!-- good, but tighten up -->
My stability results contrast, but do not necessarily contradict, the notion from competition theory that character displacement contributes to species coexistence [@Lawlor1976]. 
Competition theory typically uses a different measure of stability that considers the mutual ability of consumers with different phenotypes to invade when rare [mutual invasibility, @Chesson2000].
A shortcoming of this mutual invasibility measure, however, is that it does not allow a comparison between food webs with and without a competiting consumer. 
Such comparisons are necessary for inferring the effects of ecological character displacement, a point that has been made clear in the criteria to demonstrate character displacement [@Schluter1992; @Schluter2000].
Although the addition of consumers can decrease food-web resilience in the absence of evolution [@May1973], our results are primarily driven by an eco-evolutionary feedback between consumer evolution and resource abundances.

I expect that my theoretical predictions are robust to different assumptions about the underlying evolutionary dynamics of consumer foraging traits.
This is because models that explicitly include resource dynamics inevitably show that resource competition results in divergent character displacement, regardless of whether a quantitative genetic or adaptive dynamics model is used for consumer traits [@Lawlor1976; @Taper1985]. 
A quantitative genetic model may certainly show differences in the magnitude of character displacement, but this should not affect the qualtiative relationship I observed.  
In fact, recent work by @McPeek2017 has shown that divergence in a quantitative genetics model can lead to decreased resource abundances and unstable dynamics.
<!-- important caveat -->
I want to note though, that my conclusions may only apply to food webs with biotic resources that are nutritionally substitutable. 
It would be interesting to extend these current analyses to non-substitutable resources where convergent character displacement is expected [@Abrams1987; @Fox2008].

<!--
## Conclusions
- needs revision
- do this after I rewrite/edit the rest of the Discussion

Competition theory suggested that character displacement was a stabilizing force in ecological communities [@Lawlor1976]. 
My results show the exact opposite, when a food-web perspective is taken.
Importantly, a food-web approach permits comparisons between communities in the presence of absence of a competitor, allowing a fair comparison of the effects of character displacement on food-web dynamics.
These results appear to be quite general, as they characterize the general foraging scenario where consumers are more mobile than their resources. 
These results also suggest that the very evolutionary process that generates biodiversity, creates an ecological scenario that makes that diversity fragile to loss. 
These results highlight how evolution, when considered in a food-web context, may actually be a destabilizing force in ecological communities. 
This may generally be true of consumer evolution, as was pointed out decades ago [@Rosenzweig1963].
These results also highlight a tension between adaptive process (evolution of consumers to increase feeding efficiency) and a non-adaptive process (food-web stability).
These results suggest that in systems with evidence of ecological character displacement, why do they persist? 
These results open up a new avenue for research on ecological character displacement, one that explores in more detail the dynamical consequences of consumer evolution, rather than just static effects that assume resources are persistent and stable. 
It also suggest that actually observing evidence of ecological character displacement may be difficult ...
It also highlights how evolution of key species in a food web may make these systems more susceptible to environmental change. 
It's been argued that "inferences about the evolution of community structure cannot be based on ecological arguments about community stability or likely coexistence" [@McPeek2019].
In contrast, our results suggest that current patterns in food-web structure may be the result of non-adaptive processes selecting for more stable food-web configurations.
Our results also call for a revision of one of the criteria of character displacement, since it is clear that this process can result in different resource dynamics.
In fact, this resource similarity hypothesis has not been tested in natural systems, although it has been qualitatively inferred, and some quantitative estimates based on physical properties of systems have been measured to try and say these habitats are similar. (NEED TO CHECK SCHLUTER 2000 BOOK, THE AM NAT STUDY SUGGESTS THAT IT IS A COMMONLY FILLED CRITERIA.)
More generally, our results highlight how variation in food-web structure can alter eco-evolutionary feedbacks within communities [@McPeek2017], that also have consequences for their resilience to environmental disturbances.
This calls for food-web ecologists to more criticaly consider the role of evolutionary processes in shaping food-web structure and dynamics.
-->

@@

This work was inspired by discussions with Seth Rudman, Dolph Schluter, Ben Gilbert, and Kevin McCann.
Sally Otto provided much help in early analyses of the mathematical model. 
I would not have been able to take the theory as far as I did without her guidance and encouragement.
For funding support, I thank the University of British Columbia (Four-Year Fellowship to M.A. Barbour), NSERC (Discovery grant to Greg Crutsinger), and the Swiss National Science Foundation (grant 31003A_160671 to Jordi Bascompte).
